Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024

### 2025-02-21 products/manuscript/Manuscript.qmd
Part Two Completed
Part Two is completed I believe; Note to self on reference manager and double checking my ReadMe files

---
title: "Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

***I. Describing the Uncleaned Data***

@@

Primary data cleaning anticipated at this initial stage (this section to be rewritten later)

-   Remove all cases that were "NOT CONFIRMED," "UNCONFIRMED," or left blank in the STATUS category

-   Remove entries that contain any blanks to ensure all data analyzed later include complete profiles

-   Sort by Gender, County, and Age

@@

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

@@

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

@@

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

@@

@tbl-summarytable shows a summary of the data.

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

@@

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

@@

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

@@

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

>>>>

---
title: "Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

@@

**I. Uncleaned Data**

@@

Looked for Missing Data & General Trends

-   Identified Disease Burden Trends & Hotspot Counties

<!-- -->

-   Generated a stacked bar graph to visualize case distribution across
    regions. - Checked Age Distribution

<!-- -->

-   Verified that no cases had missing age values in the cleaned
    dataset.

<!-- -->

-   Conducted an exploratory analysis of disease counts by age group.

-   Evaluated Data Completeness Over Time

-   Extracted onset date information to track missing data trends.

-   Created a plot distinguishing between blank and non-blank STATUS
    entries over time.

-   Identified periods of high missingness, which may indicate data
    entry errors or reporting gaps.

Analyzed Timeline Between Key Case Events

-   Calculated time differences for:

    -   Onset date (DOO) to first lab test date (LABDATE1)

    -   First lab test (LABDATE1) to second lab test (LABDATE2)

    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)

    -   Summed these values to determine the total case duration. -
        Saved the processed dataset as a CSV file for further
        analysis. - Created a formatted HTML table to present extracted
        time intervals.

Addressed Data Quality Issues

-   Identified negative time values in lab test dates, indicating data
    entry errors

-   Implemented a filtering step to remove records with invalid date
    sequences

@@

@@

The data used in this analysis was obtained through my internship with
the Georgia Department of Health, in collaboration with the Centers for
Disease Control and Prevention (CDC). Specifically, it is derived from
the Nationally Notifiable Diseases Surveillance System (NNDSS) and
focuses on tick-borne illnesses. This dataset provides key insights into
the trends, distribution, and characteristics of reported cases,
supporting public health efforts in disease monitoring and prevention.

@@

The data cleaning and preprocessing workflow begins with loading
essential R packages for data handling (dplyr, tidyr, readxl), summary
visualization (skimr), and file path management (here). The raw dataset,
stored in an Excel file, is imported into R for structured processing.
To assess data quality and structure, we utilize functions such as
glimpse(), head(), and summary(), which provide an overview of column
names, data types, and potential inconsistencies like missing or
unexpected values. Additionally, skimr::skim() offers a detailed
summary, highlighting numerical distributions and missing data patterns.

The initial cleaning step removes records where the STATUS field is
either missing or not classified as "CONFIRMED," ensuring that only
officially verified cases are included. The dataset is then refined
further by retaining only cases with a confirmed "positive" test result,
excluding records with missing, blank, or ambiguous values (e.g.,
"inconclusive"). This process initially removed 15 cases that did not
meet the case definition in Georgia, leaving 95 valid records. Upon
further review, filtering criteria were adjusted to ensure all relevant
positive cases were retained while systematically excluding non-positive
values such as "Negative," "Equivocal," "Not detected," and other
irrelevant categories. This refinement resulted in a more accurate
dataset and a clearer summary table of key case information. The cleaned
dataset is saved at each stage to maintain version control and
reproducibility.

## Key Variables in Data Wrangling - Preparing for Analysis

To effectively present the cleaned dataset, key variables—including
DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are
displayed in a professionally formatted table using the kableExtra
package. This enhances readability and facilitates trend analysis.
Several visualizations are generated using ggplot2 to explore patterns
in tick-borne diseases in Georgia between 2019 and 2024: Disease
Distribution: A histogram illustrates the prevalence of different
tick-borne diseases, highlighting the most commonly reported cases.

**Rocky Mountain Spotted Fever (RMSF) Severity:**

A box plot of TITER1 values (log-transformed for better visualization)
provides insights into the variability of test results. RMSF severity
patterns are considered analogous to Alpha-Gal Syndrome, a topic of
interest, though not explicitly included in this dataset as it is not a
reportable disease in Georgia.

```{r}
library(here)
library(knitr)

# Generate correct file path
img_path <- here::here("results", "figures", "rmsf_titer_boxplot_log.png")

# Display the image
knitr::include_graphics(img_path)

```

**Most Common Tick-Borne Diseases:**

The dataset is filtered to identify the five most frequently reported
tick-borne diseases. A bar chart visualizes their case counts, offering
an overview of significant public health concerns.

```{r}
library(here)
library(knitr)

# Generate correct file path
img_path <- here::here("results", "figures", "disease_histogram.png")

# Display the image
knitr::include_graphics(img_path)
```

**Lyme Disease Analysis:**

Given its similarity to Alpha-Gal Syndrome
(which is of interest to me and the topic of my capstone), Lyme Disease
cases are analyzed separately. A count histogram compares positive,
negative, and other test outcomes, providing insights into testing
patterns and diagnosis rates.

```{r}
library(here)
library(knitr)

# Generate correct file path
img_path <- here::here("results", "figures", "lyme_result_count_histogram2.png")

# Display the image
knitr::include_graphics(img_path)

```

**Geographic Distribution:**

A grouped bar chart maps case counts across
Georgia counties, categorized by disease type. Due to the density of
information, this visualization is refined to focus on the top five
tick-borne diseases and their distribution across the 12 counties with
the highest confirmed case counts.
```{r}
library(here)
library(knitr)

# Generate correct file paths
img_paths <- c(
  here::here("results", "figures", "disease_county_plot.png"),
  here::here("results", "figures", "top5_disease_county_plot.png")
)

# Display the images
knitr::include_graphics(img_paths)
```

To analyze disease burden trends and identify hotspot counties, a
stacked bar graph is generated to visualize case distribution across
different regions.

**Age Distribution Analysis**

Age distribution is assessed as part of
the initial data cleaning process, confirming that no cases in the
cleaned dataset have missing age values. Additionally, an exploratory
analysis examines disease counts by age group to identify potential
patterns in disease prevalence.
```{r}
library(here)
library(knitr)

# Generate correct file path
img_path <- here::here("results", "figures", "disease_count_by_age.png")

# Display the image
knitr::include_graphics(img_path)
```

**Data Completeness Over Time** 
To evaluate how data completeness has
evolved, the dataset is processed to extract onset date information. A
visualization distinguishes between blank and non-blank entries in the
STATUS column over time, helping to identify periods with high
missingness. Spikes in blank values may indicate data entry errors,
system issues, or reporting gaps that should be addressed before
conducting further analysis.

```{r}
library(here)
library(knitr)

# Generate correct file path
img_path <- here::here("results", "figures", "status_completeness_plot.png")

# Display the image
knitr::include_graphics(img_path)
```

**Timeline Analysis of Key Case Events** Given the frequent
loss-to-follow-up observed in tick-borne illness reporting, an analysis
is conducted to examine time gaps between critical dates, including:

**Onset date (DOO) to first lab test date (LABDATE1)**

**First lab test (LABDATE1) to second lab test (LABDATE2)**

**Second lab test (LABDATE2) to case update date (UPDATEDATE)**

These durations are summed to determine the total case timeline for each
record. The processed dataset is saved as a CSV file for further
analysis, and a formatted HTML table is generated to present the
extracted time intervals in a structured manner. Data Quality
Adjustments Upon reviewing the time calculations, inconsistencies were
identified where negative time values appeared between initial lab
results and follow-up tests. These errors, likely due to data entry
mistakes, made it impossible to determine the correct dates. To address
this, a filtering step was added to remove erroneous records where
follow-up test dates occurred before initial lab dates.

@@

@@

@@

@@

@@
